---
title: "PT5 Rough Draft 4.7.26"
author: "Hayden and George"
editor_options:
  chunk_output_type: console
output:
  html_document:
    df_print: paged
---

# Monarch Butterflies

TODO: image of migratory map or generation cycle should go here

```{r include=FALSE}
# setup chunk

setwd("~/Documents/VIZ_SPRING/data_viz_project_spring")

library(tidyverse)
library(RColorBrewer)
library(ggplot2)
library(lubridate)
library(dplyr)
library(stringr)
library(sf)
library(plotly)
library(leaflet)
library(mapview)
library(gganimate)
library(gifski)
library(paletteer)
library(ggthemes)

# counts observed at each site over time
df = read_csv("xerces.csv", n_max = 387)

# lat and long info for each location
site_locations = read_csv("sites.csv", 
                          col_select = c("SiteID","Site Name",
                                         "Latitude","Longitude"))

# a little bit of wrangling...
df_long = pivot_longer(df, 
                       cols = !c('SITE ID','SITE NAME','COUNTY or REGION'),
                       names_to = 'time_period',
                       values_to = 'count') %>% 
          mutate(
                season = case_when(
                    str_starts(time_period, "ES") ~ "Early Season",
                    str_starts(time_period, "LS") ~ "Late Season",
                    TRUE ~ "Peak Season"),
                year = as.numeric(str_match(time_period, "[0-9]{4}")),
                time_period = as.numeric(time_period)) 

# join lat and longs to site IDs
df_long = left_join(df_long, site_locations, by=c("SITE ID" = "SiteID"))

# only observations at peak season
df_peak = df_long %>% filter(season == "Peak Season")

# observed sightings of monarchs and milkweed plants
sightings = read_csv("wmmm/monarchs_milkweed.csv", 
                     col_select = c("MilkweedObservationID","MonarchObservationID",
                                    "DateIdentified","RecordYear","State","County",
                                    "HabitatType","Threats","Latitude","Longitude"))

# categorize milkweed or monarch in one column
sightings = sightings %>% 
    mutate(
        type = case_when(
            !is.na(MilkweedObservationID) ~ "Milkweed",
            TRUE ~ "Monarch"))
```

### Monarch Butterflies have one of the most incredible migrations of any animal species. 

The eastern North American Monarch Butterfly travels a staggering 4800 km from southern Canada to northern Mexico every year. This impressive distance makes the Monarch's migration one of the farthest animal migrations, adjusting for body weight. 


### Unfortunately, the Western Monarch Butterfly population has declined significantly since 1997.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# TODO: call out 2020 count, don't show every single year

site_count_years = df_peak %>% 
    drop_na() %>%
    group_by(time_period) %>%
    summarize(total_counts = sum(count))

# counts per year displayed with a bar chart
bar_chart <- ggplot(site_count_years,
                     aes(x=time_period, y=total_counts)) +
    geom_col(fill = "Orange2") +
    labs(title = 'Western Monarch Butterflies Observed Each Year',
         x = 'Year', y = 'Count of Monarchs') +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = .5)
    )
ggplotly(bar_chart)

# counts per year displayed with a line graph
line_graph <- ggplot(site_count_years, 
                     aes(x=time_period, y=total_counts)) +
    geom_line(color="Orange2") +
    geom_point(size=0.8) +
    labs(title = 'Western Monarch Butterflies Observed Each Year', 
         x = 'Year', y = 'Count of Monarchs') +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = .5)
    )
ggplotly(line_graph)

# idk which is better


# https://www.statology.org/how-to-use-ggplot2-with-plotly-in-r-for-interactive-visualizations/
```

### At their observed peak in 1997, over 1.2 million Monarch butterflies were seen overwintering in California.

Many Western Monarchs overwinter in southern California near the coast. Hover over points to see the name of the overwintering site and how many butterflies were seen there. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_1997 = df_peak %>% filter(year == 1997) %>% drop_na()

# total_1997 = sum(df_1997$count, na.rm=TRUE)
# CA = map_data('state', 'california')

map = read_sf("north-america.geo.json")

p = ggplot(map) +
    geom_sf(
        fill = 'white',
        color = 'snow3'
    ) +
    # coord_sf(xlim=c(-125,-114), ylim=c(39,31)) +
    coord_sf(xlim=c(-127,-114), ylim=c(40,31)) +
    geom_point(
        data = df_1997, aes(x=Longitude, y=Latitude, 
                            size=count, color=count,
                            text = paste(`SITE NAME`, '\nCount:',count)),
        alpha = 0.5,
        na.rm = TRUE) +
    scale_color_gradientn(
        colors=paletteer_c("ggthemes::Red-Gold", 30)) +
    theme_void() +
    theme(panel.background = element_rect(fill='#dceced'))
    
ggplotly(p, tooltip="text")
```

```{r}
p_milk = ggplot(map) +
    geom_sf(
        fill = 'white',
        color = 'snow3'
    ) +
    # coord_sf(xlim=c(-125,-114), ylim=c(39,31)) +
    coord_sf(xlim=c(-127,-114), ylim=c(40,31)) +
    geom_point(
        data = df_peak, aes(x=Longitude, y=Latitude, 
                            size=count, color=count,
                            text = paste(`SITE NAME`, '\nCount:',count),
                            frame=year),
        alpha = 0.5) +
    scale_color_gradientn(
        colors=paletteer_c("ggthemes::Red-Gold", 30)) +
    theme_void() +
    theme(panel.background = element_rect(fill='#dceced'))
    
```


### Since then, the number of butterflies observed overwintering at each site has decreased.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p_anim = ggplot(map) +
    geom_sf(
        fill = 'white',
        color = 'snow3'
    ) +
    # coord_sf(xlim=c(-125,-114), ylim=c(39,31)) +
    coord_sf(xlim=c(-127,-114), ylim=c(40,31)) +
    geom_point(
        data = df_peak, aes(x=Longitude, y=Latitude, 
                            size=count, color=count,
                            text = paste(`SITE NAME`, '\nCount:',count),
                            frame=year),
        alpha = 0.5) +
    scale_color_gradientn(
        colors=paletteer_c("ggthemes::Red-Gold", 30)) +
    theme_void() +
    theme(panel.background = element_rect(fill='#dceced'))
    
ggplotly(p_anim, tooltip = "text") %>% 
    animation_opts(frame = 800, easing = "linear", redraw = FALSE)
```


Sources

https://xerces.org/
https://www.monarchmilkweedmapper.org/
https://www.monarchparasites.org/oe 
