---
title: "PT5 Rough Draft 4.7.26"
author: "Hayden and George"
editor_options:
  chunk_output_type: console
output:
  html_document:
    df_print: paged
---


```{r}
library(tidyverse)
library(skimr)
library(RColorBrewer)

df = read_csv("Xerces Site Counts by County.csv", n_max = 387)

distance_data_raw = read_csv("distance_per_gram.csv")

# reorder distance_data

distance_data <- distance_data_raw %>%
  arrange(`km/gram`) %>%    # First sort by val. This sort the dataframe but NOT the factor levels
  mutate(Animal = factor (Animal, levels = Animal)) 
# view(df)
# summary(df)
```

1) Monarch Butterflies have one of the most incredible migrations of any animal species. The eastern North American Monarch Butterfly travels a staggering 4800 km from southern Canada to northern Mexico every year. This impressive distance makes the Monarch's migration one of the longest animal migrations, adjusting for body weight. 

```{r}

lollipop_chart1 <- filter(distance_data, Animal != 'Brown Planthopper') %>%
  ggplot(aes(y = Animal, x = `km/gram`)) +
  geom_segment(aes(x=0, xend=`km/gram`, y=Animal, yend=Animal), color = 'gray') +
  geom_point(aes(color = Type), alpha = 0.9) +
  scale_x_log10() +
  scale_color_manual(values = c('Terrestrial' = '#4daf4a', 'Marine' = '#377eb8', 'Bird' = '#e41a1c', 'Insect' = '#ff7f00'))+
  # scale_size_continuous(range = c(1, 10)) +
  labs(title = "Notable Animal Migrations\nAdjusting for Body Weight", x = 'km/gram (log)', y = NULL) +
  theme_classic()
  

lollipop_chart1

```

2) Unfortunately, the Western Monarch Butterfly population has declined significantly since 1997.

```{r}
df_long = pivot_longer(df, 
                       cols = !c('SITE ID','SITE NAME','COUNTY or REGION'),
                       names_to = 'time_period',
                       values_to = 'count')

df_long = df_long %>% mutate(
    time_period = factor(time_period, levels = tail(names(df), -3))
  )

site_count_years = df_long %>% 
  drop_na() %>%
  group_by(time_period) %>%
  summarize(total_counts = sum(count))

site_count_years_final = df_long %>% 
  drop_na() %>%
  filter(!startsWith(as.character(time_period), "LS") & 
         !startsWith(as.character(time_period), "ES")) %>%
  group_by(time_period) %>%
  summarize(total_counts = sum(count))

bar_chart2 <- ggplot(site_count_years_final, 
                     aes(x=time_period, y=total_counts)) +
    geom_col(fill = "Orange2") +
    labs(title = 'Western Monarch Butterfly Population Over Time', 
         x = 'Year', y = 'Count of Monarchs') +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = .5)
    )

bar_chart2

```

3) As a result of the declining Monarch population, efforts to track their population has increased. In California, the number of Monarch tracking sites has steadily grown since 1997.

```{r}

df2 <- read_csv("sites_monitored.csv", n_max = 43)

midseason_sites <- df2 %>%
  filter(Type == 'Mid-Season')

#sites_sum <- df2 %>%
 # group_by(Year) %>%
  #summarise(total_sites = sum(`Number of sites monitored`))

line_chart <- ggplot(midseason_sites, aes( x = Year, y = `Number of sites monitored`)) +
  geom_line(color = 'orange') +
  geom_point() +
  labs(title = 'Monarch Butterfly Sites in California', y = 'Sites Monitored') +
  theme_classic()

line_chart

# Maybe this would be better as an infographic with little boxes for each site?
```




```{r}

# scatter_plot1 <- filter(distance_data, Animal != 'Brown Planthopper') %>%
#   ggplot(aes(x = log(`Distance (miles)`), y = log(`km/gram`), color = Type)) + # size = log(`Weight Max (grams)`)) +
#   geom_point(alpha = 0.9) +
#  # geom_text(data = filter(distance_data, Animal == 'Monarch butterfly'),aes(label = Animal)) +
#   scale_color_manual(values = c('Terrestrial' = '#4daf4a', 'Marine' = '#377eb8', 'Bird' = '#e41a1c', 'Insect' = '#ff7f00'))+
#   scale_size_continuous(range = c(1, 10)) +
#   labs(title = "Migration Distance adjusted for Weight", x = "km (log)", y = "km/grams (log)", size = "Weight (log)") +
#   theme_classic()
# 
# scatter_plot1

```

```{r}
# 
# scatter_plot2 <- filter(distance_data, Animal != 'Brown Planthopper') %>%
#   ggplot(aes(x = log(`Distance (miles)`), y = log(`km/gram`), color = Type, size = log(`Weight Max (grams)`))) +
#   geom_point(alpha = 0.9) +
# geom_text(data = filter(distance_data, Animal == 'Monarch butterfly'),aes(label = Animal)) +
#   scale_color_manual(values = c('Terrestrial' = '#4daf4a', 'Marine' = '#377eb8', 'Bird' = '#e41a1c', 'Insect' = '#ff7f00'))+
#   scale_size_continuous(range = c(1, 10)) +
#   geom_hline(yintercept = c(0)) +
#   geom_vline(xintercept = c(8.25)) +
#   labs(title = "Migration Distance adjusted for Weight", x = "Distance (log)", y = "km per gram (log)", size = "Weight (log)") +
#   theme_classic()
# 
# scatter_plot2

```



```{r}

# theme_4 <- theme_classic() +
# theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
# 
# lollipop_chart4 <- filter(distance_data, Animal != 'Brown Planthopper') %>%
#   ggplot(aes(x = Animal, y = `km/gram`)) +
#   geom_segment(aes(y=0, yend=`km/gram`, x=Animal, xend=Animal), color = 'gray') +
#   geom_point(aes(color = Type), alpha = 0.9) +
#   scale_y_log10() +
#   scale_color_manual(values = c('Terrestrial' = '#4daf4a', 'Marine' = '#377eb8', 'Bird' = '#e41a1c', 'Insect' = '#ff7f00'))+
#   # scale_size_continuous(range = c(1, 10)) +
#   labs(title = "Migration Distance adjusted for Weight", x = NULL) +
#   theme_4
#   
# 
# lollipop_chart4

```